A Recipe for Scavenging - the natural history of a behaviour

### 2016-03-31 natural history of scavenging.tex
Scavenging bats YAY!

\title{A Recipe for Scavenging - the natural history of a behaviour}
\running{Scavenging in vertebrates}

@@

\affiliations{
\item A. Kane (\url{adam.kane@ucc.ie}), University College Cork, Cooperage Building, School of Biological Earth and Environmental Sciences, Cork, Ireland.
\item K. Healy and A. Jackson, Trinity College Dublin, Department of Zoology, School of Natural Sciences, Dublin Ireland.
\item T. Guillerme, Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.
\item G. Ruxton, School of Biology, Sir Harold Mitchell Building, Greenside Place, St Andrews, KY16 9TH, United Kingdom.
}

@@

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals.
This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour.
Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, ``prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters."
He contrasts these with ``nobler kinds" of carnivores such as lions and tigers \citep{tenney1877few}.
Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact.
Without catching a carnivore in the act of killing we are left to infer how the prey was killed.
Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.
But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted.
The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms.
Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature.
Fortunately, research on scavenging is on the rise \citep{manga2006vulture}.
As a result, we are now beginning to realise the extent of this behaviour such that, ``in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{beasley2015vertebrates}.
Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses.
By recognising the difficulty in directly observing scavenging, a suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper.
Here we chart the natural history of scavenging by looking at the potential for the behaviour in dominant vertebrate groups.
%It is our aim in this review to employ these methods to gain an understanding of scavengers past and present.

@@

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts of energetically expensive activity. 
By contrast, modern reptiles are ectothermic, limiting their activity periods.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
These constraints manifests themselves in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger most if not all are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cats and the pursuit/ pounce strategies of dogs, the latter being more suited to scavenging \citep{janis2014forelimb}. 
We can (and have) use(d) these insights to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene 56-33.9 Million years ago (Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 M ya) were likely to have been ambush predators which would have counted against them taking a lot of carrion.  
%The order Carnivora saw its origins in the Middle Eocene  %TG: a phylo side note: the correct sentence would be the "The stem Carnivora saw its origins ... when it split into Caniforma and Feliforma." But I think we can leave it out because, god, it sounds pedantic!
% So we can trace efficient terrestrial movement by carnivores from this point on.

@@

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs. %TG: I will down that "dominance" term. Maybe change the sentence to "Of course the importance of terrestrial tetrapods in ecosystems predates the evolution of the dinosaurs."
It is during the Permian (298.9 - 252.17 Mya) %, almost 300 millions years ago, %TG: Being pedantic here but Permian is just *not* 300 Mya... As a side note for the dates, I say we just use the geological epochs/periods/whatever since they are the actual precise dating of the fossil occurence (and they sound better). I suggest we just add the dates in brackets to give an indication.
that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
And it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range widely, a vital component in seeking out carrion.

@@

%TG: I think it's missing a transition here: maybe something along "But moving on land as it's limitations..."
But it is in the air that we find scavengers par excellence. 
Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic}. 
We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}.
The advantage of flight can be extended further in larger species that engage in soaring instead of flapping flight, which is even cheaper still (approximately twice the basal metabolic rate) \citep{hedenstrom1993migration,spivey2014analysing}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors} and seabirds \citep{thaxter2012seabird}.
In vultures we have the best known scavengers on Earth. 
These birds consist of two convergent groups, from the old and the new world where they represent the only example of obligate vertebrate scavengers today.
The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014}.
Yet, avian flight is far older than this and originates in the Late Jurassic (163.5-145 Mya), conincident with the fossils of \textit{Archaeopteryx lithographica}, so many of these benefits would have been realised from that point on for carnivorous birds.
And vertebrate flight is much older still where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before \citep{witton2008reappraisal}.
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, look to have engaged in soaring flight \citep{witton2010size}.
%TG: this paragraph's a killer! Love it!

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge.
The bat fossil record is notoriously poor owing to their fragile skeletons so we are unable to determine if extinct species were more suited to this lifestyle \citep{eiting2009global}. %TG: again, I'm not sure it's notoriously poor! It might be poor for mammalogist standards but it's definitely a way better than birds fossil records for example!
Although it does not seem that flight is the main criterion precluding them from scavenging (see below).

Aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the cost of swimming is lower than either running or flying \citep{williams1999evolution}. 
This has led some researchers to argue for the feasability of a scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
As with the aerial and terrestrial enviornments we have evidence of facultative scavenging among extinct aquatic species.
For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark, \textit{Squalicorax} \citep{schwimmer1997scavenging}.
As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
Extant White sharks \textit{Carcharodon carcharias} too are known to feed on whale carcasses \citep{fallows2013white}. We might expect then that by combining an aquatic environment and an endothermic metabolism that marine mammals would prosper as scavengers. 
We know fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite this movement away from land the energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
This is not to say that aquatic mammalian scavengers don't exist, only that their total energetic cost is similar to an equivalent terrestrial mammal. 
%\cite{williams1999evolution} notes, ``free-ranging animals must contend with the total energetic expenditure associated with supporting basic biological functions as well as with moving the body and appendages through the environment."

>>>>

\title{A Recipe for Scavenging - the natural history of a behaviour}
%TG: nice title! I'd replace the hyphen with a comma though
\running{Scavenging in vertebrates}

@@

\affiliations\{
\item A. Kane (\url{adam.kane@ucc.ie}), University College Cork, Cooperage Building, School of Biological Earth and Environmental Sciences, Cork, Ireland.
\item K. Healy and A. Jackson, Trinity College Dublin, Department of Zoology, School of Natural Sciences, Dublin Ireland.
\item T. Guillerme, Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.
\item G. Ruxton, School of Biology, Sir Harold Mitchell Building, Greenside Place, St Andrews, KY16 9TH, United Kingdom.
}

@@

\section*{Introduction}
%TG: it might be better to reorganise the intro as:
%§1 - scavenging is understudied because it's hard to say who's a scavenger
%§2 - it's even more problematic in the fossil record but people have tried some indirect approaches
%§3 - good thing is that we can apply these indirect approaches to extant systems
%§4 - in this review we discuss some stuff.
%TG: or something along these lines, obviously flexible!
Historically, scavengers have not been viewed as the most charismatic of animals.
This may go some way to explaining the gap in our knowledge of the prevalence of this behaviour.
Consider Professor Sanborn Tenney writing in 1877 for The American Naturalist who had this to say about one well known group, ``prominent among the mammalian scavengers are the hyenas, the ugliest in their general appearance of all the flesh eaters."
He contrasts these with ``nobler kinds" of carnivores such as lions and tigers \citep{tenney1877few}.
Even aside from our own subjective biases, scavenging is a difficult behaviour to detect after the fact.
Without catching a carnivore in the act of killing we are left to infer how the prey was killed.
Some simple heuristics can inform us, for instance, in cases where the prey item was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.
But clearly, a scavenger doesn’t only feed on animals too big for it to have hunted.
The obvious lack of direct behavioural data compounds the difficulty of discerning scavenging among extinct forms.
Indeed, a single species of dinosaur notwithstanding, a synthesis describing the natural history of scavengers is absent from the literature.
%TG: I don't understand the dinosaur part of this sentence. Might be my non-native english but I thought it's worth noting for other non-english folk out there (even though I know they aren't many!).
Fortunately, research on scavenging is on the rise \citep{manga2006vulture}.
As a result, we are now beginning to realise the extent of this behaviour such that, ``in some ecosystems, vertebrates have been documented to assimilate as much as 90\% of the available carrion" \citep{beasley2015vertebrates}.
Even Tenney’s noble big cats are now known to take in a significant portion of carrion in their diet where some lion populations get over 50\% of their meat from carcasses.
By recognising the difficulty in directly observing scavenging, a suite of methods have been used to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper.
Here we chart the natural history of scavenging by looking at the potential for the behaviour in dominant vertebrate groups.
%It is our aim in this review to employ these methods to gain an understanding of scavengers past and present.

@@

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts%TG: not sure about "bouts" get it with the context and I know it makes the whole review a bit more literary and thus nicer to read but it might exclude non-native english readers. 
 of energetically expensive activity. 
By contrast, modern reptiles are ectothermic, limiting their activity periods.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
These constraints manifests themselves in aspects such as maximum sustainable speed where an equivalent mammal has a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger most (if not all) %TG: or maybe no brackets... Up to you Adam.
 are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cats and the pursuit/ pounce strategies of dogs, the latter being more suited to scavenging \citep{janis2014forelimb}. 
We can (and have) use(d) these insights to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Million years ago; Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya) were likely to have been ambush predators which would have counted against them taking a lot of carrion.  
%The order Carnivora saw its origins in the Middle Eocene  %TG: a phylo side note: the correct sentence would be the "The stem Carnivora saw its origins ... when it split into Caniforma and Feliforma." But I think we can leave it out because, god, it sounds pedantic!
% So we can trace efficient terrestrial movement by carnivores from this point on.

%TG: missing some transition here. Maybe something short like "Further back in time" or "Previous to that"...

@@

Of course, tetrapod terrestrial dominance predates the evolution of the dinosaurs. %TG: I will down that "dominance" term. Maybe change the sentence to "Of course the importance of terrestrial tetrapods in ecosystems predates the evolution of the dinosaurs."
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
And it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range widely, a vital component in seeking out carrion.

@@

%TG: I think it's missing a transition here: maybe something along "But moving on land as it's limitations..."
But it is in the air that we find scavengers \textit{par excellence}. %TG: not sure how reviewers will go with French but I like it!
Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic}. 
We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}.
The advantage of flight can be extended further in larger species that engage in soaring instead of flapping flight, which is even cheaper still \citep[pproximately twice the basal metabolic rate; ][]{hedenstrom1993migration,spivey2014analysing}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors} and seabirds \citep{thaxter2012seabird}.
In vultures we have the best known scavengers on Earth. 
These birds consist of two convergent groups, from the old and the new world where they represent the only example of obligate vertebrate scavengers today. %TG: Or "obligate vertebrate scavengers known to date." For a more emphasizing turn.

%TG: I thought the paragraph was a bit long so here's a cut for "flying scavengers in the past"
The families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014}.
Yet, avian flight is far older than this and originates in the Late Jurassic (163.5-145 Mya), conincident with the fossils of \textit{Archaeopteryx lithographica}, so many of these benefits would have been realised from that point on for carnivorous birds.
And vertebrate flight is much older still where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before \citep{witton2008reappraisal}.
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, look to have engaged in soaring flight \citep{witton2010size}.
%TG: this paragraph's a killer! Love it!

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge.
However, some fossils evidences as early as in the Eocene (56 to 33.9 Mya) suggest the presence of possible scavenging bats.
A big bat from the middle to late Eocene (56 to 33.9 Mya) bat with a robust cranio-mandibular morphology, \textit{Necromatis} (``death-eater''), could have intake carrion as a decent proportion of its diet \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.
However, this example might be peculiar to this exceptional fossil since most other fossils seems less suited to this lifestyle \citep{eiting2009global}. 
%Although it does not seem that flight is the main criterion precluding them from scavenging (see below).
%TG: Actually, maybe this example fits better in the food processing part (and maybe change the silhouettes in the picture then)?
%``The name Necromantis (``death-eater'') alludes to the diet that Weithofer assumed for this bat – carrion, but probably also small, live vertebrates, including mammals such as rodents and other bats.'' from the paper \cite{Hand_Necromantis_2012}.

Aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the cost of swimming is lower than either running or flying \citep{williams1999evolution}. 
This has led some researchers to argue for the feasibility of a scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
As with the aerial and terrestrial environments we have evidence of facultative scavenging among extinct aquatic species.
For example, the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark, \textit{Squalicorax} \citep{schwimmer1997scavenging}.
As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and mysticete whale from Peru \citep{ehret2009caught}.
Extant White sharks \textit{Carcharodon carcharias} too are known to feed on whale carcasses \citep{fallows2013white}. We might expect then that by combining an aquatic environment and an endothermic metabolism that marine mammals would prosper as scavengers. 
We know fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite this movement away from land the energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
This is not to say that aquatic mammalian scavengers don't exist, only that their total energetic cost is similar to an equivalent terrestrial mammal. 
%\cite{williams1999evolution} notes, ``free-ranging animals must contend with the total energetic expenditure associated with supporting basic biological functions as well as with moving the body and appendages through the environment."
